Simulation Tools for Small Area Estimation: Introducing the \proglang{R}-package \proglang{saeSim}

### 2015-02-12 saeSim.Rnw
update nach review 2

<<eval=FALSE>>=
library(magrittr)
colMeans(matrix(rnorm(10), ncol = 2))
rnorm(10) %>% matrix(ncol = 2) %>% colMeans
@

@@

<<eval=TRUE, echo=FALSE>>=
library(saeSim)
@

<<eval=FALSE>>=
library(saeSim)
setup1 <- sim_base_lm() %>% sim_sample(sample_number(5))
setup2 <- sim_base_lm() %>% sim_sample(sample_fraction(0.05))
@

@@

To add the area-level predictions from a Fay-Herriot model we need to define another component. The function takes a \proglang{data.frame} as input and returns the modified version. For the estimation of the EBLUP under the FH model we use the function \proglang{eblupFH} from the package \proglang{sae} \citep{molina13}. Hence we define a function named \proglang{comp\_FH} and add it to the process:

<<>>=
comp_FH <- function(dat) {
  modelFH <- sae::eblupFH(y ~ x, vardir, data = dat)
  dat$FH <- modelFH$eblup
  dat
}
setup <- setup %>% sim_comp_agg(comp_FH)
setup
@

@@

<<>>=
library(dplyr)
simResults <- sim(setup, R = 100) %>% rbind_all
simResults %>% select(idD, idR, simName, trueStat, y, FH)
@

@@

<<eval=FALSE>>=
library(ggplot2)
ggplot(ggDat, aes(x = region, y = prediction - trueStat, fill = method)) + 
  geom_boxplot() + theme(legend.position = "bottom")
ggplot(ggDat, aes(x = region, y = (prediction - trueStat)^2, fill = method)) + 
  geom_boxplot() + scale_y_log10() + theme(legend.position = "bottom")
@

<<designSimBIAS, fig = TRUE, width = 8.3, height=4, include = FALSE, echo = FALSE>>=
library(ggplot2)
ggplot(ggDat, aes(x = region, y = prediction - trueStat, fill = method)) + 
  geom_boxplot() + theme(legend.position = "bottom")
@

@@

\begin{figure}[!h]
\includegraphics[width = \textwidth]{saeSim-designSimMSE.pdf}
\caption{Monte Carlo MSE of direct vs. BHF predictor. 50 predictions for each region and estimation technique.}
\label{fig:dbMSE}
\end{figure}
\clearpage

@@

A more technical feature is a back-end for parallel computations which is a link to the \proglang{parallel} package in \proglang{R} \citep{r14}. Tools to process result data after the simulation, i.e. summaries or plotting methods, are avenues for further research. Already available are some simple plots for the simulation setups as well as a summary method to get information on the expected runtime and structure of the resulting data.

>>>>

<<eval=FALSE>>=
library("magrittr")
colMeans(matrix(rnorm(10), ncol = 2))
rnorm(10) %>% matrix(ncol = 2) %>% colMeans
@

@@

<<eval=TRUE, echo=FALSE>>=
library("saeSim")
@

<<eval=FALSE>>=
library("saeSim")
setup1 <- sim_base_lm() %>% sim_sample(sample_number(5))
setup2 <- sim_base_lm() %>% sim_sample(sample_fraction(0.05))
@

@@

To add the area-level predictions from a Fay-Herriot model we need to define another component. The function takes a \proglang{data.frame} as input and returns the modified version. For the estimation of the EBLUP under the FH model we use the function \proglang{eblupFH} from the package \proglang{sae} \citep{molina13}. To avoid naming conflicts between the dependencies of \proglang{sae} and the package \proglang{dplyr} \citep{wickham14a} we make use of the double colon operator in order to access the function without attaching the package. Hence we define a function named \proglang{comp\_FH} and add it to the process:

<<>>=
comp_FH <- function(dat) {
  modelFH <- sae::eblupFH(y ~ x, vardir, data = dat)
  dat$FH <- as.numeric(modelFH$eblup)
  dat
}

setup <- setup %>% sim_comp_agg(comp_FH)
setup
@

@@

<<>>=
library("dplyr")
simResults <- sim(setup, R = 100) %>% rbind_all
simResults %>% select(idD, idR, simName, trueStat, y, FH)
@

@@

<<eval=FALSE>>=
library("ggplot2")
ggplot(ggDat, aes(x = region, y = prediction - trueStat, fill = method)) + 
  geom_boxplot() + theme(legend.position = "bottom")
ggplot(ggDat, aes(x = region, y = (prediction - trueStat)^2, fill = method)) + 
  geom_boxplot() + scale_y_log10() + theme(legend.position = "bottom")
@

<<designSimBIAS, fig = TRUE, width = 8.3, height=4, include = FALSE, echo = FALSE>>=
library("ggplot2")
ggplot(ggDat, aes(x = region, y = prediction - trueStat, fill = method)) + 
  geom_boxplot() + theme(legend.position = "bottom")
@

@@

\begin{figure}[!h]
\includegraphics[width = \textwidth]{saeSim-designSimMSE.pdf}
\caption{Monte Carlo MSE of direct vs. BHF predictor. 50 predictions for each region and estimation technique.}
\label{fig:dbMSE}
\end{figure}

@@

A more technical feature is a back-end for parallel computations which is a link to the \proglang{parallelMap} package in \proglang{R} \citep{bischl15}. Tools to process result data after the simulation, i.e. summaries or plotting methods, are avenues for further research. Already available are some simple plots for the simulation setups as well as a summary method to get information on the expected runtime and structure of the resulting data.

### 2015-02-12 saeSim.tex
update nach review 2

\begin{Schunk}
\begin{Sinput}
> library(magrittr)
> colMeans(matrix(rnorm(10), ncol = 2))
> rnorm(10) %>% matrix(ncol = 2) %>% colMeans
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> library(saeSim)
> setup1 <- sim_base_lm() %>% sim_sample(sample_number(5))
> setup2 <- sim_base_lm() %>% sim_sample(sample_fraction(0.05))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> set.seed(1)
> setup <- base_id(nDomains = 40, nUnits = 1) %>% 
+   sim_gen_x(mean = 0, sd = 4) %>%
+   sim_gen_v(mean = 0, sd = 1)
> setup
\end{Sinput}
\begin{Soutput}
  idD          x          v
1   1 -2.5058152 -0.1645236
2   2  0.7345733 -0.2533617
3   3 -3.3425144  0.6969634
4   4  6.3811232  0.5566632
5   5  1.3180311 -0.6887557
6   6 -3.2818735 -0.7074952
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> gen_e <- function(dat) {
+   dat$vardir <- seq(0.1, 4, length.out = nrow(dat))
+   dat$e <- rnorm(nrow(dat), sd = sqrt(dat$vardir))
+   dat
+ }
> setup <- setup %>% sim_gen(gen_e)
> setup
\end{Sinput}
\begin{Soutput}
  idD          x          v vardir          e
1   1 -2.2746749 -0.5059575    0.1  0.1344285
2   2 -0.5407145  1.3430388    0.2 -0.1067262
3   3  4.7123480 -0.2145794    0.3  0.5797550
4   4 -6.0942672 -0.1795565    0.4  0.5606229
5   5  2.3757848 -0.1001907    0.5 -0.4378710
6   6  1.3318015  0.7126663    0.6  1.7088396
\end{Soutput}
\end{Schunk}

@@

To add the area-level predictions from a Fay-Herriot model we need to define another component. The function takes a \proglang{data.frame} as input and returns the modified version. For the estimation of the EBLUP under the FH model we use the function \proglang{eblupFH} from the package \proglang{sae} \citep{molina13}. Hence we define a function named \proglang{comp\_FH} and add it to the process:

\begin{Schunk}
\begin{Sinput}
> comp_FH <- function(dat) {
+   modelFH <- sae::eblupFH(y ~ x, vardir, data = dat)
+   dat$FH <- modelFH$eblup
+   dat
+ }
> setup <- setup %>% sim_comp_agg(comp_FH)
> setup
\end{Sinput}
\begin{Soutput}
  idD         x          v vardir          e         y  trueStat        FH
1   1  1.637607  0.7073107    0.1  0.1258998 104.10843 103.98253 104.06121
2   2  6.755493  1.0341077    0.2 -0.1822523 114.36284 114.54509 114.23876
3   3  6.346354  0.2234804    0.3  0.7253263 113.64151 112.91619 113.45213
4   4 -1.323631 -0.8787076    0.4 -0.4434978  96.03053  96.47403  96.45416
5   5 -9.140942  1.1629646    0.5 -0.4105563  82.47052  82.88108  82.46045
6   6  9.990646 -2.0001649    0.6 -0.7754272 117.20570 117.98113 118.08379
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> library(dplyr)
> simResults <- sim(setup, R = 100) %>% rbind_all
> simResults %>% select(idD, idR, simName, trueStat, y, FH)
\end{Sinput}
\begin{Soutput}
Source: local data frame [4,000 x 6]

@@

\begin{Schunk}
\begin{Sinput}
> setup <- simDat %>% 
+   sim_sample(sample_fraction(0.1, groupVars = "region"))
> setup
\end{Sinput}
\begin{Soutput}
      region eqIncome age agesq gender
1 Burgenland 23572.28  67  4489   male
2 Burgenland 24056.37  26   676   male
3 Burgenland 21613.73  46  2116   male
4 Burgenland 11750.30  40  1600   male
5 Burgenland  9664.08  80  6400 female
6 Burgenland 19369.91  63  3969 female
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> setup <- setup %>%
+   sim_comp_sample(comp_BHF) %>%
+   sim_comp_sample(comp_direct) %>%
+   sim_agg(agg_results)
> setup
\end{Sinput}
\begin{Soutput}
         region   direct      BHF trueStat
1    Burgenland 21933.95 21255.53 22005.42
2 Lower Austria 18885.08 19036.29 19813.37
3        Vienna 20734.08 20720.48 20395.84
4     Carinthia 19211.55 19377.21 19486.18
5        Styria 18704.75 18955.96 19335.39
6 Upper Austria 20636.94 20504.37 20517.29
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> library(ggplot2)
> ggplot(ggDat, aes(x = region, y = prediction - trueStat, fill = method)) + 
+   geom_boxplot() + theme(legend.position = "bottom")
> ggplot(ggDat, aes(x = region, y = (prediction - trueStat)^2, fill = method)) + 
+   geom_boxplot() + scale_y_log10() + theme(legend.position = "bottom")
\end{Sinput}
\end{Schunk}

>>>>

\begin{Schunk}
\begin{Sinput}
> library("magrittr")
> colMeans(matrix(rnorm(10), ncol = 2))
> rnorm(10) %>% matrix(ncol = 2) %>% colMeans
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> library("saeSim")
> setup1 <- sim_base_lm() %>% sim_sample(sample_number(5))
> setup2 <- sim_base_lm() %>% sim_sample(sample_fraction(0.05))
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> set.seed(1)
> setup <- base_id(nDomains = 40, nUnits = 1) %>% 
+   sim_gen_x(mean = 0, sd = 4) %>%
+   sim_gen_v(mean = 0, sd = 1)
> setup
\end{Sinput}
\begin{Soutput}
data.frame [40 x 3]

   idD          x          v
1    1 -2.5058152 -0.1645236
2    2  0.7345733 -0.2533617
3    3 -3.3425144  0.6969634
4    4  6.3811232  0.5566632
5    5  1.3180311 -0.6887557
6    6 -3.2818735 -0.7074952
.. ...        ...        ...
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> gen_e <- function(dat) {
+   dat$vardir <- seq(0.1, 4, length.out = nrow(dat))
+   dat$e <- rnorm(nrow(dat), sd = sqrt(dat$vardir))
+   dat
+ }
> setup <- setup %>% sim_gen(gen_e)
> setup
\end{Sinput}
\begin{Soutput}
data.frame [40 x 5]

   idD          x          v vardir          e
1    1 -2.2746749 -0.5059575    0.1  0.1344285
2    2 -0.5407145  1.3430388    0.2 -0.1067262
3    3  4.7123480 -0.2145794    0.3  0.5797550
4    4 -6.0942672 -0.1795565    0.4  0.5606229
5    5  2.3757848 -0.1001907    0.5 -0.4378710
6    6  1.3318015  0.7126663    0.6  1.7088396
.. ...        ...        ...    ...        ...
\end{Soutput}
\end{Schunk}

@@

To add the area-level predictions from a Fay-Herriot model we need to define another component. The function takes a \proglang{data.frame} as input and returns the modified version. For the estimation of the EBLUP under the FH model we use the function \proglang{eblupFH} from the package \proglang{sae} \citep{molina13}. To avoid naming conflicts between the dependencies of \proglang{sae} and the package \proglang{dplyr} \citep{wickham14a} we make use of the double colon operator in order to access the function without attaching the package. Hence we define a function named \proglang{comp\_FH} and add it to the process:

\begin{Schunk}
\begin{Sinput}
> comp_FH <- function(dat) {
+   modelFH <- sae::eblupFH(y ~ x, vardir, data = dat)
+   dat$FH <- as.numeric(modelFH$eblup)
+   dat
+ }
> setup <- setup %>% sim_comp_agg(comp_FH)
> setup
\end{Sinput}
\begin{Soutput}
data.frame [40 x 8]

   idD         x          v vardir          e         y  trueStat        FH
1    1  1.637607  0.7073107    0.1  0.1258998 104.10843 103.98253 104.06121
2    2  6.755493  1.0341077    0.2 -0.1822523 114.36284 114.54509 114.23876
3    3  6.346354  0.2234804    0.3  0.7253263 113.64151 112.91619 113.45213
4    4 -1.323631 -0.8787076    0.4 -0.4434978  96.03053  96.47403  96.45416
5    5 -9.140942  1.1629646    0.5 -0.4105563  82.47052  82.88108  82.46045
6    6  9.990646 -2.0001649    0.6 -0.7754272 117.20570 117.98113 118.08379
.. ...       ...        ...    ...        ...       ...       ...       ...
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> library("dplyr")
> simResults <- sim(setup, R = 100) %>% rbind_all
> simResults %>% select(idD, idR, simName, trueStat, y, FH)
\end{Sinput}
\begin{Soutput}
Source: local data frame [4,000 x 6]

@@

\begin{Schunk}
\begin{Sinput}
> setup <- simDat %>% 
+   sim_sample(sample_fraction(0.1, groupVars = "region"))
> setup
\end{Sinput}
\begin{Soutput}
data.frame [2,500 x 5]

       region eqIncome age agesq gender
1  Burgenland 23572.28  67  4489   male
2  Burgenland 24056.37  26   676   male
3  Burgenland 21613.73  46  2116   male
4  Burgenland 11750.30  40  1600   male
5  Burgenland  9664.08  80  6400 female
6  Burgenland 19369.91  63  3969 female
..        ...      ... ...   ...    ...
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> setup <- setup %>%
+   sim_comp_sample(comp_BHF) %>%
+   sim_comp_sample(comp_direct) %>%
+   sim_agg(agg_results)
> setup
\end{Sinput}
\begin{Soutput}
data.frame [9 x 4]

          region   direct      BHF trueStat
1     Burgenland 21933.95 21255.53 22005.42
2  Lower Austria 18885.08 19036.29 19813.37
3         Vienna 20734.08 20720.48 20395.84
4      Carinthia 19211.55 19377.21 19486.18
5         Styria 18704.75 18955.96 19335.39
6  Upper Austria 20636.94 20504.37 20517.29
..           ...      ...      ...      ...
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
> library("ggplot2")
> ggplot(ggDat, aes(x = region, y = prediction - trueStat, fill = method)) + 
+   geom_boxplot() + theme(legend.position = "bottom")
> ggplot(ggDat, aes(x = region, y = (prediction - trueStat)^2, fill = method)) + 
+   geom_boxplot() + scale_y_log10() + theme(legend.position = "bottom")
\end{Sinput}
\end{Schunk}
